### 2010-08-22 Thesis/Chapter4.tex
- Update 23 August -
Auto-review of chapters 3 and 4 + summary of chapter 4 added



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@157 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Modelling the deformation of solid objects in real-time}
\label{chap4}
\begin{shortAbstract}
A short abstract for the upcoming chapter
\end{shortAbstract}

@@

At the beginning of computer simulation, users were immersed into environments that were merely a decor. If it could virtually take the user somewhere else, a static scene quickly showed its limits. Indeed, in a real-life environment many objects are in motion, interacting with each others. Therefore, in order to increase the fidelity of simulations, elements of physics were progressively added. It could be as simple as adding motion to clouds in a flight simulator for instance. But by improving the dynamism of the scene, the simulation appears more natural to users. Over time, with the increase in computational power of computers, objects stopped being all rigid and simulators started taking laws of physics into account. In other words, they became deformable. If this took the realism a step further, demands in computational power were multiplied. Simulators eventually faced a challenge: they had to weight the desired degree of realism against the computational power at their disposal.

@@

If the reasons differ, the need for modelling the mechanical response of organs with precision remains. However, strong time constraints often limit the complexity of the modelling and affects the overall accuracy. Because of this trade-off, various approaches were proposed over the years to fit into real-time constraints and may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated usually relying on Newton's second law and (3) methods actually based on the equations of continuum mechanics.

@@

	\subsection{Free-form deformation}
The first needs to deforming solid objects came from for the field of solid modelling. The goal was to allow the designer to shape an object into the form he wishes in a similar way that clay is manipulated by a sculptor's hands. \cite{Sederberg86} introduced a technique called \emph{free-form deformation} based on the paper of \cite{Barr84}. It involves a mapping from $ \field{R}^3 $ to $ \field{R}^3 $ through a tensor product trivariate Bernstein polynomial. The deformed shape of the object is interpolated using control points as in the case of Bézier curves and surfaces patches (see Fig.~{chap4:fig-cubeFFD}). 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=6.5cm]{chapter4/cubeFFD1.pdf}
\includegraphics[width=6.6cm]{chapter4/cubeFFD2.pdf}
\end{center}
\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped. Control points are indicated by small white diamonds. The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points (right).}
\label{chap4:fig-cubeFFD}
\end{figure}
The capability of locally applying deformations makes the technique strongly analogous to sculpting with clay, which was the sought target. Although the authors do not give details on computational efficiency, their work constitutes the beginnings of deformable solid objects.

@@

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see Fig.~\ref{chap4:fig-shoes}). Another illustration of this stability is shown Fig.~\ref{chap4:fig-duck}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time.}
\label{chap4:fig-shoes}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=11cm]{chapter4/duck1.pdf} \\
\vspace{0.1cm}
\includegraphics[width=11cm]{chapter4/duck2.pdf}
\end{center}
\caption[Squeezing of a duck using Shape Matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique. Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.}
\label{chap4:fig-duck}
\end{figure}

Later, \cite{Rivers07} extented the technique to regular lattices with embedded geometry. From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh. The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions. A particule is placed at each vertex of the lattice and associated with a shape matching region. Their technique of \emph{Fast Lattice Shape Matching} allows them to compute per-region transformations in an efficient manner (see Fig.~\ref{chap4:fig-penguins}).	
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/penguins.pdf}
\end{center}
\caption[Penguins animation with Fast Lattice Shape Matching]{(\textbf{Left}) $150$ cartoon penguins deforming dynamically using Fast Lattice Shape Matching. (\textbf{Right}) Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles). The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
\label{chap4:fig-penguins}
\end{figure}

One limitation of fast lattice shape matching is that small features of a geometry yield an explosion of the runtime cost. Indeed, a small surface feature may require fine sampling in order to be deformed independently from non-adjacent material, but this fine sampling must be applied to the whole object. Therefore, \cite{Steinemann08} further improved the technique by allowing dynamic adaptive selection of levels of detail through the use of an octree-based hierarchical sampling. This adaptative sampling is illustrate with Fig.~\ref{chap4:fig-flowers}. Their technique also handles efficient topological changes.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/flowers.pdf}
\end{center}
\caption[Flowers in the wind]{Complex scene with $40$ deforming flowers. A coarse sampling is employed when the flowers are moved by the wind, for a total of $5\,680$ nodes in the scene. However, the models are dynamically refined when touched by the user, as shown in the left image.}
\label{chap4:fig-flowers}
\end{figure}

@@

	\subsection{3D Chainmail algorithm}
The \emph{3D Chainmail} was introduced by \cite{Gibson97} as a fast algorithm for deforming solid objects. Rather than carrying out complex computations on a low resolution geometry (as opposed to FEM techniques), the main idea is to take advantage of the original data resolution (of a CT scan for instance) but perform simple deformation calculation for each element. In this algorithm, each volume element is linked to its six nearest neighbours. If a link between two nodes is stretched to its limit, displacements are transferred to its neighbouring links. It works in a very similar way to a chain and is illustrated by Fig.~\ref{chap4:fig-chainmail}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=12cm]{chapter4/chainmail.pdf}
\end{center}
\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved.}
\label{chap4:fig-chainmail}
\end{figure}	
%
The chain mail algorithm is then followed by an elastic relaxation, which relaxes the shape of the deformed object approximated by the chain mail algorithm. This step consists of minimising an elastic energy measure within the object to insure that the calculated deformation is a minimum energy configuration.

@@

	\subsection{Modal analysis}
\emph{Modal analysis} was first introduced to the graphics community by \cite{Pentland89} as a fast method for approximating deformation. Modal analysis is the process of taking the non-linear description of a system, finding a good linear approximation, and then finding a coordinate system that diagonalises the linear approximation. This process transforms a complicated system of non-linear equations into a simple set of decoupled linear equations that may be individually solved analytically \citep{Hauser03}. Moreover, because each decoupled equation may be solved analytically, stability concerns due to the choice of a time integration procedure are eliminated. The linearised system of equations may be written as:
\begin{equation}
\label{chap4:linearisedEquation}
\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K} \mathbf{U} = \mathbf{R}.
\end{equation}	
where $ \mathbf{M} $ is the mass matrix, $\mathbf{D}$ is the damping matrix, $ \mathbf{K} $ is the stiffness matrix and $\mathbf{R}$ are externally applied loads. Modal decomposition refers to the process of diagonalising equation \eqref{chap4:linearisedEquation}. By using Rayleigh damping (see discussion on damping in section \ref{chap3:wordOnMatrices}), \eqref{chap4:linearisedEquation} may be rewritten as:
\begin{equation}
\label{chap4:eqRayleigh}
\mathbf{K} (\mathbf{U} + \alpha_1  \mathbf{ \dot U}) + \mathbf{M} (\alpha_2 \mathbf{ \dot U} + \mathbf{\ddot U})  = \mathbf{R},
\end{equation}	
where $ \alpha_1 $ and $ \alpha_2 $ are the Rayleigh coefficients. We denote $ \mathbf{W} $ the solution to the generalised eigenproblem $ \mathbf{K} \mathbf{x} + \lambda \mathbf{M} \mathbf{x} = 0 $ and $\boldsymbol \Lambda$ the diagonal matrix of eigenvalues, then \eqref{chap4:eqRayleigh} may be expressed as:
\begin{equation}
\boldsymbol \Lambda (\mathbf{Z} + \alpha_1  \mathbf{ \dot Z}) + (\alpha_2 \mathbf{ \dot Z} + \mathbf{\ddot Z})  = \mathbf{G},
\end{equation}
where $ \mathbf{Z} = \mathbf{W}^{-1} \mathbf{U} $ is the vector of modal coordinates and $ \mathbf{G} = \mathbf{W}^T \mathbf{U} $ is the external force vector in the modal coordinate system. Each row of this equation now corresponds to a single scalar differential equation:
\begin{equation}
\ddot z_i + (\alpha_1 \lambda_i +\alpha_2) \dot z_i + \lambda_i z_i = g_i.
\end{equation}
The analytical solutions to each equations are well known and are the following:
\begin{equation}
z_i = c_1 e^{t \omega_i^+} + c_2 e^{t \omega_i^-}
\end{equation}
where $ c_1 $ and $ c_2 $ are complex constants and $ \omega_i $ is the complex frequency given by:
\begin{equation}
\omega_i^{\pm} = \dfrac{-(\alpha_1 \lambda_i + \alpha_2) \pm \sqrt{(\alpha_1 \lambda_i +\alpha_2)^2 - 4 \lambda_i}}{2}.
\end{equation}
The main advantage of modal analysis is the possibility to model only the necessary modes. Indeed if the eigenvalue $ \lambda_i $ associated with a particular mode is large, then the force required to cause a discernible displacement of that mode will also be large. Similarly, some displacements may be too small to be detected. Removing those modes from the computation will not change the appearance of the resulting simulation while substantially reducing the computational cost.

\cite{James02} went even further by using graphics hardware to reduce CPU costs to a minimum. In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see Fig.~\ref{chap4:fig-modalAnalysisLaparo}). The technique is fairly efficient as it allows precomputations of the vibration modes. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter4/modalAnalysisLaparo.pdf}
\end{center}
\caption[Applications of modal analysis]{(\textbf{Left}) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis. (\textbf{Right}) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations. } 
\label{chap4:fig-modalAnalysisLaparo}
\end{figure}

@@

	\subsection{Mass-spring model}
A classic technique for modelling a deformable object is the mass-spring system. In this method, the geometry is described by a network of masses connected together by springs. The force $ F $ exerted onto each mass by the spring may be computed with:
\begin{equation}
\label{chap4:massSpring1}
F_{\text{spring}} = - k x,
\end{equation}
where $ k $ is the spring stiffness constant, $ x $ represents the stretch of the spring from its rest position. From our experience, we know that a mass attached to a single spring has a tendency to oscillate. If no appropriate measure is taken, a network of springs will a forciori oscillate as well. To prevent the system from oscillating, a damper is also added between each couple of masses. The force created by the damper acts to reduce the velocity and may be expressed as:
\begin{equation}
\label{chap4:massSpring2}
F_{\text{damper}} = - d v
\end{equation}
where $ d $ is the damping constant and $ v $ the derivative of $ x $ with respect to time. The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations. Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{a} $ yields the standard differential equation for a mass-spring system:
\begin{equation}
\label{chap4:massSpring3}
m \ddot x = - d \dot x - k x.
\end{equation}
For complex systems with more than one spring attached to each mass, the right-hand side of the equation above becomes more complex as it needs to be repeated and adjusted for each spring. The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}). Dividing the total force by the mass gives the current acceleration of each mass (by \eqref{chap4:massSpring3}). This equation must be numerically solved for each mass, either by explicit or implicit methods. A discussion of integration methods for mass-spring models can be found in \cite{Shinya05}.

@@

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in Fig.~\ref{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behavior of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=8cm]{chapter4/undesiredAnisotropy.pdf}
\end{center}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system (left) shows signs of anisotropy, the hexahedral mesh (right) with springs aligned along the gravity does not.} 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

>>>>

\chapter{Modelling the deformation of solid objects in real-time}
\label{chap4}
\begin{shortAbstract}
Modelling the deformation of solid organs is a crucial problem in medical simulation. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation. Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course. Over time, along with the increase in computational power, methods became more elaborate and less approximative. These different approaches may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods actually based on the equations of continuum mechanics. In this chapter, we will review the most classic methods applied to modelling of solid objects with a particular emphasis on medical simulation. 
\end{shortAbstract}

@@

At the beginning of computer simulation, users were immersed into environments that were merely a decor. If it could virtually take the user somewhere else, a static scene quickly showed its limits. Indeed, in a real-life environment many objects are in motion, interacting with each others. Therefore, in order to increase the fidelity of simulations, elements of physics were progressively added. It could be as simple as adding motion to clouds in a flight simulator for instance. But by improving the dynamism of the scene, the simulation appears more natural to users. Over time, with the increase of computers' computational power, objects stopped being all rigid and simulators started taking laws of physics into account. In other words, objects became deformable. If this took the realism a step further, the demands in computational power were multiplied. Simulators eventually faced a challenge: they had to weight the desired degree of realism against the computational power at their disposal.

@@

If the reasons differ, the need for modelling the mechanical response of organs with precision remains. However, strong time constraints often limit the complexity of the modelling and affects the overall accuracy. Because of this trade-off, various approaches were proposed over the years to fit into real-time constraints and may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated (usually relying on Newton's second law) and (3) methods actually based on the equations of continuum mechanics.

@@

	\subsection{Free-form deformation}
The first needs to deforming solid objects came from for the field of solid modelling. The goal was to allow the designer to shape an object into the form he wishes in a similar way that clay is manipulated by a sculptor's hands. \cite{Sederberg86} introduced a technique called \emph{free-form deformation} based on the paper of \cite{Barr84}. It involves a mapping from $ \field{R}^3 $ to $ \field{R}^3 $ through a tensor product trivariate Bernstein polynomial. The deformed shape of the object is interpolated using control points as in the case of Bézier curves and surface patches (see \fig{chap4:fig-cubeFFD}). 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=6.5cm]{chapter4/cubeFFD1.pdf}
\includegraphics[width=6.6cm]{chapter4/cubeFFD2.pdf}
\end{center}
\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped \textbf{(left)}. Control points are indicated by small white diamonds. The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points \textbf{(right)}.}
\label{chap4:fig-cubeFFD}
\end{figure}
The capability of locally applying deformations makes the technique strongly analogous to sculpting with clay, which was the sought target. Although the authors do not give details on computational efficiency, their work constitutes the beginnings of deformable solid objects.

@@

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see \fig{chap4:fig-shoes}). Another illustration of this stability is shown \fig{chap4:fig-duck}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time.}
\label{chap4:fig-shoes}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=11cm]{chapter4/duck1.pdf} \\
\vspace{0.1cm}
\includegraphics[width=11cm]{chapter4/duck2.pdf}
\end{center}
\caption[Squeezing of a duck using shape matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique. Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.}
\label{chap4:fig-duck}
\end{figure}

Later, \cite{Rivers07} extented the technique to regular lattices with embedded geometry. From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh. The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions. A particule is placed at each vertex of the lattice and associated with a shape matching region. Their technique of \emph{fast lattice shape matching} allows them to compute per-region transformations in an efficient manner (see \fig{chap4:fig-penguins}).	
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/penguins.pdf}
\end{center}
\caption[Penguins animation with fast lattice shape matching]{\textbf{(Left)} $150$ cartoon penguins deforming dynamically using fast lattice shape matching\textbf{(Right)}. Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles). The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
\label{chap4:fig-penguins}
\end{figure}

One limitation of fast lattice shape matching is that small features of a geometry yield an explosion of the runtime cost. Indeed, a small surface feature may require fine sampling in order to be deformed independently from non-adjacent material, but this fine sampling must be applied to the whole object. Therefore, \cite{Steinemann08} further improved the technique by allowing dynamic adaptive selection of levels of detail through the use of an octree-based hierarchical sampling. This adaptative sampling is illustrate with \fig{chap4:fig-flowers}. Their technique also handles efficient topological changes.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/flowers.pdf}
\end{center}
\caption[Flowers in the wind]{Complex scene with $40$ deforming flowers. A coarse sampling is employed when the flowers are moved by the wind, for a total of $5\,680$ nodes in the scene. However, the models are dynamically refined when touched by the user, as shown in the right image.}
\label{chap4:fig-flowers}
\end{figure}

@@

	\subsection{3D Chainmail algorithm}
The \emph{3D Chainmail} was introduced by \cite{Gibson97} as a fast algorithm for deforming solid objects. Rather than carrying out complex computations on a low resolution geometry (as opposed to FEM techniques), the main idea is to take advantage of the original data resolution (of a CT scan for instance) but perform simple deformation calculation for each element. In this algorithm, each volume element is linked to its six nearest neighbours. If a link between two nodes is stretched to its limit, displacements are transferred to its neighbouring links. It works in a very similar way to a chain and is illustrated by \fig{chap4:fig-chainmail}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=12cm]{chapter4/chainmail.pdf}
\end{center}
\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved.}
\label{chap4:fig-chainmail}
\end{figure}	
%
The chain mail algorithm is then followed by an elastic relaxation, which relaxes the shape of the deformed object approximated by the chain mail algorithm. This step consists of minimising an elastic energy measure within the object to insure that the calculated deformation is a minimum energy configuration.

@@

	\subsection{Modal analysis}
\emph{Modal analysis} was first introduced to the graphics community by \cite{Pentland89} as a fast method for approximating deformation. Modal analysis is the process of taking the non-linear description of a system, finding a good linear approximation, and then finding a coordinate system that diagonalises the linear approximation. This process transforms a complicated system of non-linear equations into a simple set of decoupled linear equations that may be individually solved analytically \citep{Hauser03}. Moreover, because each decoupled equation may be solved analytically, stability concerns due to the choice of a time integration procedure are eliminated. The linearised system of equations may be written as:
\begin{equation}
\label{chap4:linearisedEquation}
\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K} \mathbf{U} = \mathbf{R}.
\end{equation}	
where $ \mathbf{M} $ is the mass matrix, $\mathbf{D}$ is the damping matrix, $ \mathbf{K} $ is the stiffness matrix and $\mathbf{R}$ are the externally applied loads. Modal decomposition refers to the process of diagonalising equation \eqref{chap4:linearisedEquation}. By using Rayleigh damping (see discussion on damping in section \ref{chap3:wordOnMatrices}), \eqref{chap4:linearisedEquation} may be rewritten as:
\begin{equation}
\label{chap4:eqRayleigh}
\mathbf{K} (\mathbf{U} + \alpha_1  \mathbf{ \dot U}) + \mathbf{M} (\alpha_2 \mathbf{ \dot U} + \mathbf{\ddot U})  = \mathbf{R},
\end{equation}	
where $ \alpha_1 $ and $ \alpha_2 $ are the Rayleigh coefficients. We denote $ \mathbf{W} $ the solution to the generalised eigenproblem $ \mathbf{K} \mathbf{x} + \lambda \mathbf{M} \mathbf{x} = 0 $ and $\boldsymbol \Lambda$ the diagonal matrix of eigenvalues, then \eqref{chap4:eqRayleigh} may be expressed as:
\begin{equation}
\boldsymbol \Lambda (\mathbf{Z} + \alpha_1  \mathbf{ \dot Z}) + (\alpha_2 \mathbf{ \dot Z} + \mathbf{\ddot Z})  = \mathbf{G},
\end{equation}
where $ \mathbf{Z} = \mathbf{W}^{-1} \mathbf{U} $ is the vector of modal coordinates and $ \mathbf{G} = \mathbf{W}^T \mathbf{U} $ is the external force vector in the modal coordinate system. Each row of this equation now corresponds to a single scalar differential equation:
\begin{equation}
\ddot z_i + (\alpha_1 \lambda_i +\alpha_2) \dot z_i + \lambda_i z_i = g_i.
\end{equation}
The analytical solutions to each equations are well known and are the following:
\begin{equation}
z_i = c_1 e^{t \omega_i^+} + c_2 e^{t \omega_i^-}
\end{equation}
where $ c_1 $ and $ c_2 $ are complex constants and $ \omega_i $ is the complex frequency given by:
\begin{equation}
\omega_i^{\pm} = \dfrac{-(\alpha_1 \lambda_i + \alpha_2) \pm \sqrt{(\alpha_1 \lambda_i +\alpha_2)^2 - 4 \lambda_i}}{2}.
\end{equation}
The main advantage of modal analysis is the possibility to model only the necessary modes. Indeed if the eigenvalue $ \lambda_i $ associated with a particular mode is large, then the force required to cause a discernible displacement of that mode will also be large. Similarly, some displacements may be too small to be detected. Removing those modes from the computation will not change the appearance of the resulting simulation while substantially reducing the computational cost.

\cite{James02} went even further by using graphics hardware to reduce CPU costs to a minimum. In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see \fig{chap4:fig-modalAnalysisLaparo}). The technique is fairly efficient as it allows precomputations of the vibration modes. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13cm]{chapter4/modalAnalysisLaparo.pdf}
\end{center}
\caption[Applications of modal analysis]{(\textbf{Left}) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis. (\textbf{Right}) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations. } 
\label{chap4:fig-modalAnalysisLaparo}
\end{figure}

@@

	\subsection{Mass-spring model}
A classic technique for modelling a deformable object is the mass-spring system. In this method, the geometry is described by a network of masses connected together by springs. The force $ F $ exerted onto each mass by the spring may be computed with:
\begin{equation}
\label{chap4:massSpring1}
F_{\text{spring}} = - k x,
\end{equation}
where $ k $ is the spring stiffness constant, $ x $ represents the stretch of the spring from its rest position. From our experience, we know that a mass attached to a single spring has a tendency to oscillate. If no appropriate measure is taken, a network of springs will a forciori oscillate as well. To prevent the system from oscillating, a damper is also added between each couple of masses. The force created by the damper acts to reduce the velocity and may be expressed as:
\begin{equation}
\label{chap4:massSpring2}
F_{\text{damper}} = - d v
\end{equation}
where $ d $ is the damping constant and $ v $ the derivative of $ x $ with respect to time. The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations. Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{a} $ yields the standard differential equation for a mass-spring system:
\begin{equation}
\label{chap4:massSpring3}
m \ddot x = - d \dot x - k x.
\end{equation}
For complex systems with more than one spring attached to each mass, the right-hand side of the equation \eqref{chap4:massSpring3} becomes more complex as it needs to be repeated and adjusted for each spring. The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}). Dividing the total force by the mass gives the current acceleration of each mass (by \eqref{chap4:massSpring3}). This equation must be numerically solved for each mass, either by explicit or implicit methods. A discussion of integration methods for mass-spring models can be found in \cite{Shinya05}.

@@

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in \fig{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behavior of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=8cm]{chapter4/undesiredAnisotropy.pdf}
\end{center}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system \textbf{(left)} shows signs of anisotropy, the hexahedral mesh \textbf{(right)} with springs aligned along the gravity does not.} 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

A mass-spring model may also be enhanced to feature viscoelasticity. Indeed, \cite{Stiles72} proposed a viscoelastic mass-spring system to test hypotheses on the mechanical response of muscles. Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a mass-spring system. They use a large number of particules to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics. \cite{Basafa10} went even further and extended a mass-spring system to simulate non-linear viscoelastic deformations of soft tissue for laparoscopic surgery. They tune their parameters by a simple optimisation procedure to fit the mechanical response obtained on a set of experimental data.
